Theory about Normal Basis on $F(2^k)$

### 2016-12-11 dissertation/recha.tex
dissertation: committee review version

\chapter{Word-Level Traversal of Finite State Machines using Algebraic Geometry}
\label{ch:reacha}
Reachability analysis is a basic component of sequential circuit verification, especially for 
formal equivalence checking and model checking techniques. Concretely, in modern synthesis 
tools, in order to improve various performance indicators such as latency, clock skew or power
density, sequential optimization techniques such as retiming \cite{retiming}, scan logic \cite{scan},
sequential power optimization \cite{mathur2009power} and clock-gating techniques \cite{clockgating} are applied.
% some major refactoring and attachments are made on original designs. 
These modifications may introduce, bugs, errors or malfunctions to the original logic and cause problems. 
Based on traditional localized simulation or formal verification method ({\it e.g.} equivalence checking), 
designers are reluctant to make aggressive optimization
since the malfunctions are considered as ``faults" in the circuits.
However, if the circuit behavior is carefully investigated, it may become evident that 
those ``faults" may never be activated during a restricted execution starting 
from legal initial states and with legal inputs. Thus we will call those ``faults" as 
{\it spurious faults} (false negatives), since they will not affect the circuit's normal behavior.

@@

\section{Motivation}
Before introducing the details of our approach, there are a few questions to ask: what is 
the benefit of execute FSM traversal at word-level? How can algebraic geometry make this happen?
The answers can be found in this section, as a statement of the motivation of our research.

\subsection{FSM Traversal Algorithms}
Sequential circuits are modeled as FSMs, which can be implemented as graphs. Thus a 
graph-traversal based algorithm is created to analysis the reachable states \cite{coudert2003unified}.
A traversal algorithm using the concept of implicit state enumeration is proposed \cite{cho1993redundancy}.
Concretely, the algorithm is given in Algorithm \ref{alg:BFS}:

@@

For each element in $\F_{2^3}$, we write down the truth table as follows:
\begin{table}[H]
\label{tab:truthtable}
\centering{\small
\begin{tabular}{|c|ccc|c|}
\hline
$\{a_2a_1a_0\}\in\mathbb{B}^3$   & $A\in \F_{2^3}$ &$\rightarrow$& $\{z_2z_1z_0\}\in\mathbb{B}^3$  & $Z\in \F_{2^3}$ \\
\hline
000  &0 &$\rightarrow$&000 & 0 \\
001  &1 &$\rightarrow$&001 & 1 \\
010  &$\alpha$ & $\rightarrow$ & 111& $\alpha^2 + \alpha + 1$ \\
011  &$\alpha + 1$ &$\rightarrow$& 111 & $\alpha^2 + \alpha + 1$ \\
100  &$\alpha^2$ &$\rightarrow$& 101 &  $\alpha^2+1$ \\
101  &$\alpha^2 + 1$ &$\rightarrow$&011 & $\alpha+1$ \\
110  &$\alpha^2 + \alpha$&$\rightarrow$& 101 &$\alpha^2 + 1$ \\
111  &$\alpha^2 + \alpha + 1$ &$\rightarrow$& 101 &$\alpha^2 + 1$\\
\hline
\end {tabular}
}
\caption{Truth table for mappings in $\mathbb{B}^3$ and $\F_{2^3}$}
\end{table}
Now our objective is to abstract a function over finite field $\F_{2^3}$ in word-level variables, {\it i.e.} 
$Z = \Func(A)$. Recall Lagrange's interpolation formula:
\begin{equation}
\label{eqn:Lagrange}
\Func(x) =  \sum_{k=1}^N \left[ \prod_{(0\leq j \leq k-1),(j\neq i)}\frac{x-x_j}{x_i-x_j} \cdot y_k \right]
\end{equation}
The geometric meaning of Lagrange's interpolation in real algebra is: given $N$ dots with coordinates $(x_i,y_i)$,
they can always be fitted into a polynomial function with at most $N-1$ degree, and that function can be 
written in the form of Equation \ref{eqn:Lagrange}. In this example, although defined in a Galois field instead of 
the real number field, the essential concept of Lagrange's interpolation remains the same. 
We can get 8 points in the affine space:
\begin{align*}
\text{Generic form}&: (a_2\alpha^2+a_1\alpha+a_0,z_2\alpha^2+z_1\alpha+z_0) \gets (A,Z) \\
\text{Point }1&: (0, 0) \gets (000,000) \\
\text{Point }2&: (1,1) \gets (001,001) \\
\text{Point }3&:  (\alpha,\alpha^2 + \alpha + 1) \gets (010,111)\\
\text{Point }4&:  (\alpha + 1,\alpha^2 + \alpha + 1)\gets (011,111) \\
\text{Point }5&:  (\alpha^2,\alpha^2+1) \gets (100,101)\\
\text{Point }6&:  (\alpha^2 + 1,\alpha+1)\gets (101,011)\\
\text{Point }7&:  (\alpha^2 + \alpha,\alpha^2 + 1)\gets (110,101)\\
\text{Point }8&:  (\alpha^2 + \alpha + 1,\alpha^2 + 1)\gets (111,101)
\end{align*}
Substitute 8 $(x_i,y_i)$ pairs in Equation \ref{eqn:Lagrange} with these 8 points in $\F_{2^3}$.
The result is a polynomial function with degree no greater than 7:
\begin{align*}
Z =& \Func(A) \\
=& (\alpha^2+\alpha+1) A^7 + (\alpha^2+1) A^6 + \alpha A^5 + (\alpha+1) A^4 \\
&+ (\alpha^2+\alpha+1)A^3+ (\alpha^2+1)A
\end{align*}

@@

\begin{Proof}
Since $\Fkk[x]$ is a principal ideal domain, $ J = \langle g\rangle$
for some polynomial $ g \in \Fkk[x]$. Let $d = gcd(g, x^{2^k} -
x)$. So, $g = dg_1 , x^{2^k} - x = df_1$, with $gcd(f_1 , g_1 ) =
1$. Then $J_0 : J = \langle f_1 \rangle$ by Lemma \ref{lem:gcd}.

@@

Since $x^{2^k} = x$, we see that either $d(x) = 0$ or $f_1 (x) =
0$. Considering $g(x) \neq 0$, we can assert that $d(x) \neq 0$. In
conclusion, $f_1 (x) = 0$ and $x \in V(f_1 )$.

@@

Now we make an assumption that $x \in V(g)$. Then $g(x) = 0 =
d(x)g_1(x)$ which means either $d(x) = 0$ or $g_1 (x) = 0$.

If $g_1 (x) = 0$, then since $f_1 (x) = 0$ we get that $f_1 , g_1$
share a root. This contradicts the fact that $gcd(f_1 , g_1 ) = 1$.

On the other hand, if $d(x) = 0$, then since $f_1 (x) = 0$ and
$x^{2^k} - x = df_1$, we get that $x^{2^k} - x$ has a double root. 
But this is impossible since the derivative of $x^{2^k} - x$ is $-1$.

So, $x \notin V(g)$ and this concludes the proof.
\end{Proof}

@@

\section{Problem Setup and Formulation}
\begin{Problem}
We use the notions from Figure \ref{fig:seqmodel}(a) in this setup. 
\begin{enumerate}[{1)}] 
\item The circuit is modeled over $\Fkk = \F_2[x]\pmod{p(x)}, p(\alpha) = 0$, where $k$ is the number of flip-flops,
	the PS variables $\{s_0,dots,s_{k-1}\}$, NS variables $\{t_0,\dots,t_{k-1}\}$.
\item Denote $S$ as the PS word-level by the following polynomial:
$$f_S: S = s_0+s_1\alpha+s_2\alpha^2+\cdots+s_{k-1}\alpha^{k-1}$$
Similarly we define NS word-level variable $T$ by:
$$f_T: T = t_0+t_1\alpha+t_2\alpha^2+\cdots+t_{k-1}\alpha^{k-1}$$
\item Impose ATO for sequential circuits which is 
$$LEX: \text{ all bit-level variables in any order }>S>T$$
\item Write polynomials $f_1,\dots,f_s$ for each gate, and construct ideal describing the circuit:
$$J_{ckt} = \langle f_1,\dots,f_s,f_S,f_T\rangle$$
as well as the ideal with vanishing polynomials:
$$J_0 = \langle \dots,x_i^2-x_i,\dots,S^{2^k}-S,T^{2^k}-T\rangle$$
where $x_i$ corresponds to bit-level variables denoting wires in the circuit.
\item Compute GB with ATO for $G$. Then obtain the projection of variety only on NS variable $T$, by
$$G\cap \Fkk[T]$$
\end{enumerate}
\end{Problem}

\IncMargin{1em}
\begin{algorithm}[hbt]
\SetAlgoNoLine
\Indm
 \KwIn{The circuit's characteristic polynomial ideal $J_{ckt}$,
   initial state polynomial $\F(S)$, and LEX term order: bit-level
   variables $x, s, t$ $>$ PS word $S$ $>$ NS word $T$}
\Indp

  $from^0 = reached = \F(S)$\;
  \Repeat{$\langle new^i\rangle == \langle 1\rangle$}
  {
  	$i \gets i + 1$\;
  	$G \gets$GB($\langle J_{ckt},J_v, from^{i-1}\rangle$)\;
  	\tcc{Compute Gr\"obner basis with elimination term order: $T$ smallest}
	$to^i \gets G\cap \Fkk[T]$\;
	\tcc{There will be a univariate polynomial in $G$ denoting the
          set of next states in word-level variable $T$}
	$\langle new^i\rangle \gets \langle to^i\rangle + (\langle T^{2^k}-T\rangle:\langle reached\rangle)$\;
	\tcc{Use quotient of ideals to attain complement of reached states, then use sum of ideals to attain an intersection with next state}
  	$\langle reached\rangle \gets \langle reached\rangle \cdot \langle new^i\rangle$\;
  	\tcc{Use product of ideals to attain a union of newly reached states and formerly reached states}
	$from^i \gets new^i(S\setminus T)$\;
	\tcc{Start a new iteration by replacing variable $T$ in newly reached states with current state variable $S$}
  }
  \tcc{Loop until a fixpoint reached: newly reached state is empty}
\Return{$\langle reached\rangle$}
\caption {Algebraic Geometry based FSM Traversal}\label{alg:univa}
\end{algorithm}
\DecMargin{1em}

@@

\IncMargin{1em}
\begin{algorithm}[hbt]
\SetAlgoNoLine
\Indm
 \KwIn{Input-output circuit characteristic polynomial ideal $J_{ckt}$, initial state polynomial $\Func(S)$}
 \KwOut{Final reachable states represented by polynomial $\mathcal G(T)$}
\Indp

@@

\subsection{Primary Inputs Partition}
Using above techniques we can get a remainder polynomial with only word-level PS/NS variables. However in most 
cases, the number of bit-level PIs will be too large for the last-step Gr\"obner basis computation. 
Therefore it is necessary to convert bit-level PIs to word-level PI variables.

@@

\section{Concluding Remarks} \label{sec:conc}
This dissertation has presented a new approach to perform reachability
analysis of finite state machines at the word-level. This is achieved
by modeling the transition relations and sets of states by way of
polynomials over finite fields $\Fkk$, where $k$ represents the size
of the state register bits. Subsequently using the concepts of
elimination ideals, Gr\"obner bases, and quotients of ideals, we show
that the set of reachable states can be encoded, canonically, as the
variety of a univariate polynomial. This polynomial is computed using
the Gr\"obner basis algorithm {\it w.r.t.} an elimination term
order. Experiments are conducted with a few FSMs that validate the
concept of word-level FSM traversal using algebraic geometry.

>>>>

\chapter{Word-Level Traversal of Finite State Machines using Algebraic Geometry}
\label{ch:reacha}
Reachability analysis is a basic component of sequential circuit verification, especially for 
formal equivalence checking and model checking techniques. Concretely, in modern synthesis 
tools, in order to improve various performance indicators such as latency, clock skew or power
density, sequential optimization techniques such as retiming \cite{retiming}, scan logic \cite{scan},
sequential power optimization \cite{mathur2009power} and clock-gating techniques \cite{clockgating} are applied.
% some major refactoring and attachments are made on original designs. 
These modifications may introduce bugs, errors or malfunctions to the original logic and cause problems. 
Based on traditional localized simulation or formal verification method ({\it e.g.} equivalence checking), 
designers are reluctant to make aggressive optimization
since the malfunctions are considered as ``faults" in the circuits.
However, if the circuit behavior is carefully investigated, it may become evident that 
those ``faults" may never be activated during a restricted execution starting 
from legal initial states and with legal inputs. Thus we will call those ``faults" as 
{\it spurious faults} (false negatives), since they will not affect the circuit's normal behavior.

@@

\section{Motivation}
Before introducing the details of our approach, there are a few questions to ask: what is 
the benefit of executing FSM traversal at word-level? How can algebraic geometry make this happen?
The answers can be found in this section, as a statement of the motivation of our research.

\subsection{FSM Traversal Algorithms}
Sequential circuits are modeled as FSMs, which can be implemented as graphs. Thus a 
graph-traversal based algorithm is created to analyze the reachable states \cite{coudert2003unified}.
A traversal algorithm using the concept of implicit state enumeration is proposed \cite{cho1993redundancy}.
Concretely, the algorithm is given in Algorithm \ref{alg:BFS}:

@@

For each element in $\F_{2^3}$, we write down the truth table as follows:
\begin{table}[H]
\label{tab:truthtable}
\centering{\small
\begin{tabular}{|c|ccc|c|}
\hline
$\{a_2a_1a_0\}\in\mathbb{B}^3$   & $A\in \F_{2^3}$ &$\rightarrow$& $\{z_2z_1z_0\}\in\mathbb{B}^3$  & $Z\in \F_{2^3}$ \\
\hline
000  &0 &$\rightarrow$&000 & 0 \\
001  &1 &$\rightarrow$&001 & 1 \\
010  &$\alpha$ & $\rightarrow$ & 111& $\alpha^2 + \alpha + 1$ \\
011  &$\alpha + 1$ &$\rightarrow$& 111 & $\alpha^2 + \alpha + 1$ \\
100  &$\alpha^2$ &$\rightarrow$& 101 &  $\alpha^2+1$ \\
101  &$\alpha^2 + 1$ &$\rightarrow$&011 & $\alpha+1$ \\
110  &$\alpha^2 + \alpha$&$\rightarrow$& 101 &$\alpha^2 + 1$ \\
111  &$\alpha^2 + \alpha + 1$ &$\rightarrow$& 101 &$\alpha^2 + 1$\\
\hline
\end {tabular}
}
\caption{Truth table for mappings in $\mathbb{B}^3$ and $\F_{2^3}$}
\end{table}
Now our objective is to abstract a function over finite field $\F_{2^3}$ in word-level variables, {\it i.e.} 
$Z = \Func(A)$. Recall Lagrange's interpolation formula:
\begin{equation}
\label{eqn:Lagrange}
\Func(x) =  \sum_{k=1}^N \left[ \prod_{(0\leq j \leq k-1),(j\neq i)}\frac{x-x_j}{x_i-x_j} \cdot y_k \right]
\end{equation}
The geometric meaning of Lagrange's interpolation in real algebra is: given $N$ points with coordinates $(x_i,y_i)$,
they can always be fitted into a polynomial function with at most $N-1$ degree, and that function can be 
written in the form of Equation \ref{eqn:Lagrange}. In this example, although defined in a Galois field instead of 
the real number field, the essential concept of Lagrange's interpolation remains the same. 
We can get 8 points in the affine space:
\begin{align*}
\text{Generic form}&: (a_2\alpha^2+a_1\alpha+a_0,z_2\alpha^2+z_1\alpha+z_0) \gets (A,Z) \\
\text{Point }1&: (0, 0) \gets (000,000) \\
\text{Point }2&: (1,1) \gets (001,001) \\
\text{Point }3&:  (\alpha,\alpha^2 + \alpha + 1) \gets (010,111)\\
\text{Point }4&:  (\alpha + 1,\alpha^2 + \alpha + 1)\gets (011,111) \\
\text{Point }5&:  (\alpha^2,\alpha^2+1) \gets (100,101)\\
\text{Point }6&:  (\alpha^2 + 1,\alpha+1)\gets (101,011)\\
\text{Point }7&:  (\alpha^2 + \alpha,\alpha^2 + 1)\gets (110,101)\\
\text{Point }8&:  (\alpha^2 + \alpha + 1,\alpha^2 + 1)\gets (111,101)
\end{align*}
Substitute 8 $(x_i,y_i)$ pairs in Equation \ref{eqn:Lagrange} with these 8 points in $\F_{2^3}$.
The result is a polynomial function with degree no greater than 7:
\begin{align*}
Z =& \Func(A) \\
=& (\alpha^2+\alpha+1) A^7 + (\alpha^2+1) A^6 + \alpha A^5 + (\alpha+1) A^4 \\
&+ (\alpha^2+\alpha+1)A^3+ (\alpha^2+1)A
\end{align*}

@@

\begin{Proof}
Since $\Fkk[x]$ is a principal ideal domain, $ J = \langle g(x)\rangle$
for some polynomial $ g(x) \in \Fkk[x]$. Let $h(x) = gcd(g(x), x^{2^k} -
x)$. So, $g(x) = h(x)g_1(x) , x^{2^k} - x = h(x)f_1(x)$, with $gcd(f_1(x) , g_1(x) ) =
1$. Then $J_0 : J = \langle f_1(x) \rangle$ by Lemma \ref{lem:gcd}.

@@

Since $x^{2^k} = x$, we see that either $h(x) = 0$ or $f_1 (x) =
0$. Considering $g(x) \neq 0$, we can assert that $h(x) \neq 0$. In
conclusion, $f_1 (x) = 0$ and $x \in V(f_1 )$.

@@

Now we make an assumption that $x \in V(g)$. Then $g(x) = 0 =
d(x)g_1(x)$ which means either $h(x) = 0$ or $g_1 (x) = 0$.

If $g_1 (x) = 0$, then since $f_1 (x) = 0$ we get that $f_1(x) , g_1(x)$
share a root. This contradicts the fact that $gcd(f_1(x) , g_1(x) ) = 1$.

On the other hand, if $h(x) = 0$, then since $f_1 (x) = 0$ and
$x^{2^k} - x = d(x)f_1(x)$, we get that $x^{2^k} - x$ has a double root. 
But this is impossible since the derivative of $x^{2^k} - x$ is $-1$.

So, $x \notin V(g(x))$ and this concludes the proof.
\end{Proof}

@@

\section{Problem Setup and Formulation}
\begin{Problem}
We use the notions from Figure \ref{fig:seqmodel}(a) in this setup. 
\begin{enumerate}[{1)}] 
\item The circuit is modeled over $\Fkk = \F_2[x]\pmod{p(x)}, p(\alpha) = 0$, where $k$ is the number of flip-flops,
	the PS variables $\{s_0,\dots,s_{k-1}\}$, NS variables $\{t_0,\dots,t_{k-1}\}$.
\item Denote $S$ as the PS word-level by the following polynomial:
$$f_S: S = s_0+s_1\alpha+s_2\alpha^2+\cdots+s_{k-1}\alpha^{k-1}$$
Similarly we define NS word-level variable $T$ by:
$$f_T: T = t_0+t_1\alpha+t_2\alpha^2+\cdots+t_{k-1}\alpha^{k-1}$$
\item Impose ATO for sequential circuits which is 
$$LEX: \text{ all bit-level variables in any order }>S>T$$
\item Write polynomials $f_1,\dots,f_s$ for each gate, and construct ideal describing the circuit:
$$J_{ckt} = \langle f_1,\dots,f_s,f_S,f_T\rangle$$
as well as the ideal with vanishing polynomials:
$$J_0 = \langle \dots,x_i^2-x_i,\dots,S^{2^k}-S,T^{2^k}-T\rangle$$
where $x_i$ corresponds to bit-level variables denoting wires in the circuit.
\item Compute GB with ATO for $G$. Then obtain the projection of variety only on NS variable $T$, by
$$G\cap \Fkk[T]$$
\end{enumerate}
\end{Problem}

\IncMargin{1em}
\begin{algorithm}[hbt]
\SetAlgoNoLine
\LinesNumbered
\Indm
 \KwIn{The circuit's characteristic polynomial ideal $J_{ckt}$,
   initial state polynomial $\F(S)$, and LEX term order: bit-level
   variables $x, s, t$ $>$ PS word $S$ $>$ NS word $T$}
\Indp

  $from^0 = reached = \Func(S)$\;
  \Repeat{$\langle new^i\rangle == \langle 1\rangle$}
  {
  	$i \gets i + 1$\;
  	$G \gets$GB($\langle J_{ckt},J_v, from^{i-1}\rangle$)\;
  	\tcc{Compute Gr\"obner basis with elimination term order: $T$ smallest}
	$to^i \gets G\cap \Fkk[T]$\;
	\tcc{There will be a univariate polynomial in $G$ denoting the
          set of next states in word-level variable $T$}
	$\langle new^i\rangle \gets \langle to^i\rangle + (\langle T^{2^k}-T\rangle:\langle reached\rangle)$\;
	\tcc{Use quotient of ideals to attain complement of reached states, then use sum of ideals to attain an intersection with next state}
  	$\langle reached\rangle \gets \langle reached\rangle \cdot \langle new^i\rangle$\;
  	\tcc{Use product of ideals to attain a union of newly reached states and formerly reached states}
	$from^i \gets new^i(S\setminus T)$\;
	\tcc{Start a new iteration by replacing variable $T$ in newly reached states with current state variable $S$}
  }
  \tcc{Loop until a fixpoint reached: newly reached state is empty}
\Return{$\langle reached\rangle$}
\caption {Algebraic Geometry based FSM Traversal}\label{alg:univa}
\end{algorithm}
\DecMargin{1em}

@@

\IncMargin{1em}
\begin{algorithm}[hbt]
\SetAlgoNoLine
\LinesNumbered
\Indm
 \KwIn{Input-output circuit characteristic polynomial ideal $J_{ckt}$, initial state polynomial $\Func(S)$}
 \KwOut{Final reachable states represented by polynomial $\mathcal G(T)$}
\Indp

@@

\subsection{Primary Inputs Partitioning}
Using above techniques we can get a remainder polynomial with only word-level PS/NS variables. However in most 
cases, the number of bit-level PIs will be too large for the last-step Gr\"obner basis computation. 
Therefore it is necessary to convert bit-level PIs to word-level PI variables.

@@

\section{Concluding Remarks}
This chapter has presented a new approach to perform reachability
analysis of finite state machines at the word-level. This is achieved
by modeling the transition relations and sets of states by way of
polynomials over finite fields $\Fkk$, where $k$ represents the size
of the state register bits. Subsequently using the concepts of
elimination ideals, Gr\"obner bases, and quotients of ideals, we show
that the set of reachable states can be encoded, canonically, as the
variety of a univariate polynomial. This polynomial is computed using
the Gr\"obner basis algorithm {\it w.r.t.} an elimination term
order. Experiments are conducted with a few FSMs that validate the
concept of word-level FSM traversal using algebraic geometry.

### 2016-12-11 dissertation/unsat.tex
dissertation: committee review version

Boolean satisfiability (SAT) problem is the basis of 
most decidable decision problems, as well as the basis of many formal verification techniques. 
In this chapter, we discuss a special topic branching out from SAT theory. It is about a situation 
when SAT problems give negative answer, which are called unsatisfiability (UNSAT) problems. Within 
a set of constrains ({\it e.g.} clauses, formulas or polynomials) which is unsatisfiable, sometimes 
it is worthy to explore the reasons for UNSAT. From the execution of the GB algorithm, 
an auxiliary structure can be obtained to help explore the reasons causing UNSAT. This chapter introduces the 
details about the motivation, mechanism and implementation.

@@

\subsection{Preliminaries of SAT/UNSAT Theory}
Boolean satisfiability (SAT) problem is a fundamental problem in computer science.
In the following part we define the terminology related to SAT problem \cite{anaICCAD}.

>>>>

The Boolean satisfiability (SAT) problem is the basis of 
most decidable decision problems, as well as the basis of many formal verification techniques. 
In this chapter, we discuss a special topic branching out from SAT theory. It is about a situation 
when SAT problems give negative answer, which are called unsatisfiability (UNSAT) problems. Within 
a set of constrains ({\it e.g.} clauses, formulas or polynomials) which is unsatisfiable, sometimes 
it is worthy to explore the reasons for UNSAT. From the execution of the GB algorithm, 
an auxiliary structure can be obtained to help explore the reasons causing UNSAT. This chapter introduces the 
details about the motivation, mechanism and implementation.

@@

\subsection{Preliminaries of SAT/UNSAT Theory}
The Boolean satisfiability (SAT) problem is a fundamental problem in computer science.
In the following part we define the terminology related to SAT problem \cite{anaICCAD}.
